Algorithm explanation

### 2013-04-27 article/algo.tex
Changes of Horn graph structure * For supporting disjunctive Horn clauses

The Horn clause solving algorithm receives an input problem in Horn
graph. A Horn graph is a labelled directed graph $G=(V,E,\varphi)$,
where V is a set of vertices and E is a set of labelled
edges. $\varphi: V \rightarrow L$ is a labelling function for
vertices, where L is a set of elements of a predicate variable $P(x_1,
..., x_n)$ or a linear inequality $a_1 x_1 + \cdots + a_n x_n \leq b$.
An edge from a vertex $u$ to $v$ is expressed as a 3-tuple
$(u,v,\theta)$ where $\theta$ is a finite map from variables to
variables.

The meaning $\llbracket G \rrbracket $ of a Horn graph $G$ is a set of
Horn clauses $\mathcal{HC}_v$ for all vertices $v \in V$.

\begin{align*}
\mathcal{HC}_v & = \left( \bigwedge_{(u,v,\theta) \in E} \theta \varphi(u) \right) \Longrightarrow \varphi(v) \\
\llbracket G \rrbracket & = \bigwedge_{v \in V} \mathcal{HC}_v
\end{align*}

@@

Given an acyclic Horn graph (or called a Horn DAG), we define the
relation $u <^+ v$ as follows.
\[ u <^+ v \Longleftrightarrow \exists \theta. (u,v,\theta) \in E \]
By the assumption that $G$ is acyclic, $<^+$ is a well-founded
relation. We assume without loss of generality that $V$ contains the
greatest element $v_\bot$ with respect to $<^+$; otherwise we can
always implicitly assume $\varphi(v_\bot) = 0 \leq 0$.

For each $v \in V$, we define
$\Delta(v) = (\mathbf{a}_v \mathbf{x} \leq b_v, C_v)$
, where
\begin{itemize}
\item $\mathbf{a}_v \mathbf{x} \leq b_v$ is a linear inequality that
  may contain coefficient variables
\item $C_v$ is a set of linear expressions on coefficient variables.
\end{itemize}
Intuitively, the pair $(\mathbf{a} \mathbf{x} \leq b, C)$ denotes the
set of inequalities:

\begin{align*}
\left\lbrace
 \mathbf{a} \mathbf{x} \leq b \middle|
 \exists \left( FV(C)
  \setminus (\mathbf{a} \cup \left\lbrace b \right\rbrace
 \right). C
\right\rbrace
\end{align*}

@@

\begin{align*}
C_v = & \bigcup_{u <^+ v} C_u
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \nexists u. u <^+ v \\
\left\lbrace
 \mathbf{a}_v = \sum_{u <^+ v} \mathbf{a}_u,
 b_v \geq \sum_{u <^+ v} b_u
\right\rbrace
& \mbox{otherwise}
\end{cases}
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \varphi(v) = P(\mathbf{x}) \\
\left\lbrace
 \lambda_v \geq 0, \mathbf{a}_v = \lambda_v \mathbf{a}_0,
 b_v = \lambda_v b_0
\right\rbrace
& \mbox{if } \varphi(v) = \mathbf{a}_0 \mathbf{x} \leq b_0
\end{cases}
\end{align*}

@@

\begin{quote}
If $\models \sigma(C_{v_\bot})$, then $\rho$ defined by
\begin{align*}
 \rho = \left\lbrace
  \left( P, \sigma(\mathbf{a}_v \mathbf{x} \leq b_v) \right) \middle|
  \forall v \in V. \varphi(v) = P(\mathbf{x})
 \right\rbrace
\end{align*}
is a solution of $G$.
\end{quote}

@@

\begin{align*}
C_v = & \bigcup_{\substack{u <^+ v \\ u \not\in \hat V}} C_u \cup
\bigcup_{\substack{u <^+ v \\ u \in \hat V}} \textsc{Copy}_v (C_u)
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \nexists u. u <^+ v \\
\left\lbrace
 \mathbf{a}_{v,\emptyset} =
  \sum_{\substack{u <^+ v \\ u \not\in \hat V}} \mathbf{a}_{u,U} +
  \sum_{\substack{u <^+ v \\ u \in \hat V}} \mathbf{a}_{u,U \cup \left\lbrace v \right\rbrace},
 b_{v,\emptyset} \geq
  \sum_{\substack{u <^+ v \\ u \not\in \hat V}} b_{u,U} +
  \sum_{\substack{u <^+ v \\ u \in \hat V}} b_{u,U \cup \left\lbrace v \right\rbrace}
\right\rbrace
& \mbox{otherwise}
\end{cases}
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \varphi(v) = P(\mathbf{x}) \\
\left\lbrace
 \lambda_v \geq 0, \mathbf{a}_{v,\emptyset} = \lambda_v \mathbf{a}_0,
 b_{v,\emptyset} = \lambda_v b_0
\right\rbrace
& \mbox{if } \varphi(v) = \mathbf{a}_0 \mathbf{x} \leq b_0
\end{cases}
\end{align*}

@@

\begin{align*}
 \rho = \left\lbrace
  \left( P, \bigwedge_S \sigma(\mathbf{a}_{v,S} \mathbf{x} \leq b_{v,S}) \right) \middle|
  \forall v \in V. \varphi(v) = P(\mathbf{x})
 \right\rbrace
\end{align*}

>>>>

The Horn clause solving algorithm receives an input problem in Horn
graph. A Horn graph is a labelled directed graph $G=(V,E,\varphi)$.
\begin{itemize}
\item $V$ is a union of two sets of vertices; Horn term vertices $V_T$
  and arrow vertices $V_\Rightarrow$. Each vertex $u \in
  V_\Rightarrow$ have eactly one outgoing edge to some $v \in V_T$,
  denoted as $succ(u)$.
\item $E$ is a set of labelled edges, which are expressed in the form
  $(u,v,\theta)$ where $\theta$ is a finite map from variables to
  variables.
\item $\varphi: V_T \rightarrow L$ is a labelling function for Horn
  term vertices, where L is a set of elements of a predicate variable
  in the form $P(x_1, ..., x_{arity(P)})$ or a linear inequality $a_1
  x_1 + \cdots + a_n x_n \leq b$.
\end{itemize}

There are two kinds of edges in a Horn graph;
\begin{itemize}
\item for $u \in V_T, v \in V_\Rightarrow$, an edge takes a form
  $(u,v,\theta)$, or,
\item for $u \in V_\Rightarrow, v \in V_T$, an edge takes a form
  $(u,v,\emptyset)$, which the mapping is empty.
\end{itemize}

The meaning $\llbracket G \rrbracket $ of a Horn graph $G$ is a set of
Horn clauses $\mathcal{HC}_v$ for all vertices $v \in V_\Rightarrow$.

\begin{align*}
\mathcal{HC}_v & = \left( \bigwedge_{(u,v,\theta) \in E} \theta \varphi(u) \right) \Longrightarrow \varphi(succ(v)) \\
\llbracket G \rrbracket & = \bigwedge_{v \in V} \mathcal{HC}_v
\end{align*}

We implicitly assume each $\mathcal{HC}_v$ is universally quantified
by its all free variables $FV(\mathcal{HC}_v)$.

Note that we allow multiple appearance of Horn clauses with the same
predicate variable on their right-hand side.

@@

Given an acyclic Horn graph (or called a Horn DAG), we define the
relation $u <^+ v$ over $V_T$ as follows.
\[ u <^+ v \Longleftrightarrow \exists a, \theta; \left\lbrace{ (u,a,\theta), (a,v,\emptyset) \right\rbrace \subseteq E \]
By the assumption that $G$ is acyclic, $<^+$ is a well-founded
relation. We assume without loss of generality that $V_T$ contains the
greatest element $v_\bot$ with respect to $<^+$; otherwise we can
always implicitly assume $\varphi(v_\bot) = 0 \leq 0$ such that $\forall v \in V_\bot; \exists v_a; \left\lbrace (v, v_a, \emptyset), (v_a, v_\bot, \emptyset) \right\rbrace \subseteq E $.

For each $v \in V_T$, we define
$\Delta(v) = (\mathbf{a}_v \mathbf{x} \leq b_v, C_v)$
, where
\begin{itemize}
\item $\mathbf{a}_v \mathbf{x} \leq b_v$ is a linear inequality that
  may contain coefficient variables
\item $C_v$ is a set of linear expressions on coefficient variables.
\end{itemize}
Intuitively, the pair $(\mathbf{a} \mathbf{x} \leq b, C)$ denotes the
set of inequalities:

\begin{align*}
\left\lbrace
 \mathbf{a} \mathbf{x} \leq b \middle|
 \exists \left( FV(C)
  \setminus (\mathbf{a} \cup \left\lbrace b \right\rbrace
 \right); C
\right\rbrace
\end{align*}

@@

\begin{align*}
C_v = & \bigcup_{u <^+ v} C_u
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \nexists u; u <^+ v \\
\left\lbrace
 \mathbf{a}_v = \sum_{u <^+ v} \mathbf{a}_u,
 b_v \geq \sum_{u <^+ v} b_u
\right\rbrace
& \mbox{otherwise}
\end{cases}
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \varphi(v) = P(\mathbf{x}) \\
\left\lbrace
 \lambda_v \geq 0, \mathbf{a}_v = \lambda_v \mathbf{a}_0,
 b_v = \lambda_v b_0
\right\rbrace
& \mbox{if } \varphi(v) = \mathbf{a}_0 \mathbf{x} \leq b_0
\end{cases}
\end{align*}

@@

\begin{quote}
If $\models \sigma(C_{v_\bot})$, then $\rho$ defined by
\begin{align*}
 \rho = \left\lbrace
  \left( P, \sigma(\mathbf{a}_v \mathbf{x} \leq b_v) \right) \middle|
  \forall v \in V; \varphi(v) = P(\mathbf{x})
 \right\rbrace
\end{align*}
is a solution of $G$.
\end{quote}

@@

\begin{align*}
C_v = & \bigcup_{\substack{u <^+ v \\ u \not\in \hat V}} C_u \cup
\bigcup_{\substack{u <^+ v \\ u \in \hat V}} \textsc{Copy}_v (C_u)
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \nexists u; u <^+ v \\
\left\lbrace
 \mathbf{a}_{v,\emptyset} =
  \sum_{\substack{u <^+ v \\ u \not\in \hat V}} \mathbf{a}_{u,U} +
  \sum_{\substack{u <^+ v \\ u \in \hat V}} \mathbf{a}_{u,U \cup \left\lbrace v \right\rbrace},
 b_{v,\emptyset} \geq
  \sum_{\substack{u <^+ v \\ u \not\in \hat V}} b_{u,U} +
  \sum_{\substack{u <^+ v \\ u \in \hat V}} b_{u,U \cup \left\lbrace v \right\rbrace}
\right\rbrace
& \mbox{otherwise}
\end{cases}
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \varphi(v) = P(\mathbf{x}) \\
\left\lbrace
 \lambda_v \geq 0, \mathbf{a}_{v,\emptyset} = \lambda_v \mathbf{a}_0,
 b_{v,\emptyset} = \lambda_v b_0
\right\rbrace
& \mbox{if } \varphi(v) = \mathbf{a}_0 \mathbf{x} \leq b_0
\end{cases}
\end{align*}

@@

\begin{align*}
 \rho = \left\lbrace
  \left( P, \bigwedge_S \sigma(\mathbf{a}_{v,S} \mathbf{x} \leq b_{v,S}) \right) \middle|
  \forall v \in V; \varphi(v) = P(\mathbf{x})
 \right\rbrace
\end{align*}
